WSClean: an implementation of a fast, generic wide-field imager

### 2014-02-09 articles/2014/Article-WSClean/wsclean.tex
Better fits

\begin{table*}
\caption{Predicted imaging times for various surveys, based on multi-frequency synthesis with 5 major iterations observing for 1 hour at ZA=20\degree.  Predictions are for $w$-projection in CASA, $w$-stacking with WSClean and $w$-stacked 12$\times$5-min snapshots with WSClean, with approximately equal accuracy.}
\begin{tabular}{lrrrrrrrrrr|rrr}
 Survey   & $\lambda$ & FWHM  & Beams & Ant & Res & $\max |\Delta \mathbf{x}|$ & $\max \Delta z$ & int. t & BW & $\Delta \nu$ & $w$-proj & $w$-stack & $w$-snaps \\
          &           &       &       &     &     & (km)                       &                 & (s)    & (MHz) &  (kHz) &  &  & \\
\hline
 GLEAM    &  2 m   & 24.7\degree& 1    & 128 & 2'   & 2.9 & 5~m & 2 & 32 & 40     &14h 02m & 1h 52 m & 1h 23m \\
 EMU      & 20 cm  & 1\degree   & 36   &  36 & 10''  & 6 &  0~m &10 & 300 & 20    &19h 49m &14h 12 m & 14h 13m\\
 MSSS low &  5 m   & 9.8\degree & 5    &  20 & 100'' & 5 &      &10 & 16 & 16     & 2h 30m & 0h 11 m & 0h 11m\\
 MSSS high&  2 m   & 3.8\degree & 5    &  40 & 120'' & 5 &      &10 & 16 & 16     & 2h 31m & 0h 38 m & 0h 39m\\
 VLSS     & 4.1 m  & 14\degree  & 1    &  27 &  80'' &11.1& 0~m &10 & 1.56 & 12.2 & 0h 32m & 0h 12 m & 0h 02m\\
\end{tabular}
\end{table*}

\subsection{Optimal snapshot time}
The computing time of snapshot imaging can be expressed in either of the $w$-correction techniques. Snapshot imaging effectively removes the dependency on ZA from $w_e$. Given the snapshot time $\Delta t_\textrm{snapshot}$ and the total time $\Delta t_\textrm{total}$, the total time cost of inversion becomes
\begin{equation}
t(\Delta t_\textrm{snapshot}) = \frac{\Delta t_\textrm{total}}{\Delta t_\textrm{snapshot}} t_\textrm{imaging}(w'_e,N'_\textrm{vis},N_\textrm{freq},N_\textrm{pix}),
\end{equation}
with
\begin{equation}
w'_e = \frac{1}{\lambda}\max \Delta z \left(1.0 - \cos(\frac{1}{2}(\textrm{FOV} + \angle t_\textrm{snapshot})\right)
\end{equation}
and $N'_\textrm{vis} = \frac{N_\textrm{vis}\Delta t_\textrm{total}}{t_\textrm{snapshot}}$. This function can be minimized to find the optimal snapshot time. For WSClean with MWA parameters, we find this function does not reach its minimum within 24~h, suggesting that the optimal snapshot time is as long as possible. This implies that the change in in performance due to higher $w$-values because of Earth rotation during an individual snapshot is less significant than the cost of the additional FFTs. For $w$-projection in CASA we find an optimal snapshot time of 122~seconds.

\subsection{Optimal $w$-project/$w$-stacking combination}
The $w$-projection and $w$-stacking algorithms can be combined by gridding onto several $w$-layers to which large $w$-terms are corrected after the FFTs, and using a $w$-correcting kernel for making small $w$-corrections before the FFTs. This allows using small $w$-kernels and limits at the same time the number of FFTs and required memory. The AWImager has been applied like this, which let to better results compared to $w$-projection without $w$-stacking \citep{awimager-2013}. For this scenario, Eq.~\eqref{eq:casa-computing-cost} can be used to determine the optimal number of $w$-layers, or more generally the optimal distance between layers, by assuming that the cost of calculating a single $w$-layer equals the cost of imaging the data set with correspondingly smaller $w_e$ value and smaller $N_\textrm{vis}$. If $\Delta w$ is the distance between $w$-layers, then 
\begin{equation}
t(\Delta w) = \frac{w_e}{\Delta w} t_\textrm{CASA}(w'_e,N'_\textrm{vis},N_\textrm{freq},N_\textrm{pix}),
\end{equation}
with $N'_\textrm{vis} = \frac{N_\textrm{vis}\Delta w}{w_e}$ and $w'_e = \Delta w.

@@

@@

\section{Optimal gridding kernel size} \label{sec:gridding-appendix}
\begin{figure}
\begin{center}
\includegraphics[width=8cm]{img/benchmark-kernelsize/kernel}
\caption{Effect of antialiasing kernel size on performance.}
\label{fig:timing-kernelsize}
\end{center}
\end{figure}
As discussed in \S\ref{sec:gridding}, WSClean uses a Kaiser-Bessel window against aliasing. The type of window function has no effect on the gridding performance, but the size of the kernel affects gridding performance quadratically. Fig.~\ref{fig:timing-kernelsize} shows that this quadratic relation becomes significant for kernel sizes $\gtrapprox 10$. Decreasing the kernel size to values below $7$ has little effect on performance, because for small kernels the visibility-reading rate is lower than the gridding rate. We have not noticed much benefit of larger kernels except in rare cases where a bright source lies just outside the imaged field of view. Therefore, WSClean uses a default of 7 pixels for the gridding kernel. It can be increased when necessary. Of course, different hardware might give slightly different results because of different reading and calculation performance.

>>>>

TODO say something about error of analytical function vs measurements.

\begin{table*}
\caption{Predicted imaging times for various surveys, based on multi-frequency synthesis with 5 major iterations observing for 1 hour at ZA=20\degree.  Predictions are for $w$-projection in CASA, $w$-stacking with WSClean and $w$-stacked 12$\times$5-min snapshots with WSClean, with approximately equal accuracy.}
\begin{tabular}{lrrrrrrrrrr|rrr}
 Configuration&$\lambda$&FWHM  & Beams & Ant & Res & $\max |\Delta \mathbf{x}|$ & $\max \Delta z$ & int. t & BW & $\Delta \nu$ & $w$-proj & $w$-stack & $w$-snaps \\
          & (m)       &       &       &     &     & (km)                       & (m)             & (s)    & (MHz) &  (kHz) &  &  & \\
\hline
 GLEAM    &  2   & 24.7\degree& 1    & 128 & 2'   & 2.9 & 5 & 2 & 32 & 40     &14h 02m & 1h 52 m & 1h 23m \\
 EMU      & 0.2  & 1\degree   & 30   &  36 & 10''  & 6 &  0 &10 & 300 & 20    &19h 49m &14h 12 m & 14h 13m\\
 MSSS low &  5   & 9.8\degree & 5    &  20 & 100'' & 5 &    &10 & 16 & 16     & 2h 30m & 0h 11 m & 0h 11m\\
 MSSS high&  2   & 3.8\degree & 5    &  40 & 120'' & 5 &    &10 & 16 & 16     & 2h 31m & 0h 38 m & 0h 39m\\
 AARTFAAC &  5   & 45\degree  & 1    & 288 &  20'  & 0.3& 1 & 1 &  7 & 24     & & & \\
 VLSS     & 4.1  & 14\degree  & 1    &  27 &  80'' &11.1& 0 &10 & 1.56 & 12.2 & 0h 32m & 0h 12 m & 0h 02m\\
 SKA-P1 AA inner&3& 1.2\degree& 480  &  35 & 140'' &2.5& 2 &1.2 & 380 & 1     &       &         &      \\
 SKA-P1 AA mid&3 & 1.2\degree & 480  &  50 & 3.5'' &100& 50 &1.2 & 380 & 1     &       &         &      \\
\end{tabular}
\end{table*}

\subsection{Optimal snapshot duration}
Snapshot imaging can in theory be implemented with either $w$-projection or $w$-stacking. Using the derived formula, we can estimate the cost of making snapshots with both techniques. Snapshot imaging effectively removes the dependency on ZA from $w_e$. Given the snapshot duration $\Delta \tau_\textrm{snapshot}$ and the total integration time $\Delta \tau_\textrm{total}$, the total time cost of inversion becomes
\begin{equation} \label{eq:snapshot-cost}
t_\textrm{snapshot}(\Delta \tau_\textrm{snapshot}) = \frac{\Delta \tau_\textrm{total}}{\Delta \tau_\textrm{snapshot}} t_\textrm{imaging}(w'_e,N'_\textrm{vis},N_\textrm{freq},N_\textrm{pix}),
\end{equation}
with
\begin{equation}
w'_e = \frac{1}{\lambda}\max \Delta z \left(1.0 - \cos(\frac{1}{2}(\textrm{FOV} + \omega_E \tau_\textrm{snapshot})\right),
\end{equation}
$\omega_E$ the rotational speed of the Earth and $N'_\textrm{vis} = N_\textrm{vis}\frac{\Delta \tau_\textrm{total}}{\tau_\textrm{snapshot}}$. We have excluded the cost of phase shifting the visibilities and gridding. The cost for regridding with simple nearest neighbour or bilinair interpolation is indeed negligable, although more accurate interpolation (e.g. Lanczos interpolation) can be expensive. Also, our current implementation of the phase-changing program does take a non-negligable time, but this implementation is not optimized and can in theory be implemented in a preprocessing pipeline or on the fly during imaging.

Function $t_\textrm{snapshot}$ can be minimized to find the optimal snapshot duration. For WSClean with MWA parameters, we find that this function decreases but no minimum is reached for $\Delta \tau_\textrm{snapshot}<24$~h. Therefore, from a performance perspective, the snapshot duration should be as large as possible. For $w$-projection in CASA we find an optimal snapshot duration of 122~s for MWA observations.

\subsection{Combination of $w$-projection and $w$-stacking}
To combine the $w$-projection and $w$-stacking algorithms, small $w$-corrections are made before the FFTs using a $w$-correcting kernel and large $w$-terms are corrected after the FFTs by gridding onto several $w$-layers. This allows using small $w$-kernels and limits at the same time the number of FFTs and required memory. The AWImager has been applied like this, which let to better results compared to $w$-projection without $w$-stacking \citep{awimager-2013}. For this scenario, Eq.~\eqref{eq:casa-computing-cost} can be used to determine the optimal number of $w$-layers, or more generally the optimal distance between layers, by assuming that the cost of calculating a single $w$-layer equals the cost of imaging the data set with correspondingly smaller $w_e$ value and smaller $N_\textrm{vis}$. If $\Delta w$ is the distance between $w$-layers, then 
\begin{equation}
t(\Delta w) = \frac{w_e}{\Delta w} t_\textrm{CASA}(w'_e,N'_\textrm{vis},N_\textrm{freq},N_\textrm{pix}),
\end{equation}
with $N'_\textrm{vis} = N_\textrm{vis}\frac{\Delta w}{w_e}$ and $w'_e = \Delta w$.

@@

TODO discuss fact that w-projection needs layers for multi-threading.

@@

\section{Optimal gridding kernel size} \label{sec:gridding-appendix}
\begin{figure}
\begin{center}
\includegraphics[width=8cm]{img/benchmark-kernelsize/kernel}
\caption{Effect of antialiasing-kernel size on performance.}
\label{fig:timing-kernelsize}
\end{center}
\end{figure}
As discussed in \S\ref{sec:gridding}, WSClean uses a Kaiser-Bessel window against aliasing. The type of window function has no effect on the gridding performance, but the size of the kernel affects gridding performance quadratically. Fig.~\ref{fig:timing-kernelsize} shows that this quadratic relation becomes significant for kernel sizes $\gtrapprox 10$. Decreasing the kernel size to values below $7$ has little effect on performance, because for small kernels the visibility-reading rate is lower than the gridding rate. We have not noticed much benefit of larger kernels except in rare cases where a bright source lies just outside the imaged field of view. Therefore, WSClean uses a default of 7 pixels for the gridding kernel. It can be increased when necessary. Of course, different hardware might give slightly different results because of different reading and calculation performance.
